\textbf{Matching the forecast horizon with the relevant spatial and temporal processes and data sources}

### 2020-04-29 doc/manuscript.tex
removed red ink

\noindent \textbf{Abstract:}   Most phenomenological, statistical models used to generate ecological forecasts take either a time-series approach, based on long-term data from one location, or a space-for-time approach, based on data describing spatial patterns across environmental gradients. \new{However, the magnitude and even the sign of environment-response relationships detected using these two approaches often differs, leading to contrasting predictions about responses to future environmental change.} Here we consider how the forecast horizon determines whether more accurate predictions come from the time-series approach, the space-for-time approach, or a combination of the two. We use two simulated case studies to show that forecasts for short and long forecast horizons need to focus on different ecological processes, which are reflected in different kinds of data. In the short-term, dynamics reflect initial conditions and fast processes such as birth and death, and the time-series approach makes the best predictions. In the long-term, dynamics reflect the additional influence of slower processes such as evolutionary and ecological selection, colonization and extinction, which the space-for-time approach can effectively capture. At intermediate time-scales, a weighted average of the two approaches shows promise. However, making this weighted model operational will require new research to predict the rate at which slow processes begin to influence dynamics.

@@

Although both time-series and space-for-time approaches are widely used, there has been little discussion of their advantages and disadvantages for guiding policy decisions or advancing our understanding of ecological dynamics \citep{harris_forecasting_2018,Renwick2018}. \new{Such a discussion is overdue given that these two approaches are likely to make very different predictions about ecological responses to future environmental change. For example, average forage production in the shortgrass steppe increases rapidly with increasing mean annual precipitation, but production is much less sensitive to interannual variation in precipitation at any one location \citep{lauenroth_long-term_1992}. The space-for-time approach would predict large changes in production in response to a future change in precipitation, while the time-series approach would predict smaller changes. Bird species' abundances in the United Kindgom also show different responses to spatial vs. temporal variation in weather covariates. Not only are abundances typically more sensitive to spatial variation in the covariates, but in many cases the temporal response takes the opposite sign \citep{Oedekoven2017}. Whenever the sign or magnitude of relationships based on time-series and space-for-time approaches differ, so will the resulting forecasts.}

@@

Here we use two simulated case studies to \new{1) demonstrate why time-series and space-for-time approaches can make different predictions, 2) propose} that the best model-building approaches for ecological forecasting 
may depend on the time horizon of the forecast, and 3) explore how time-series and space-for-time approaches might be 
combined via weighted averaging to make better forecasts at intermediate time scales. The first case study
focuses on how interspecific interactions affect the population dynamics of a focal species, and the second focuses on
an eco-evolutionary scenario. Our analyses show that: 
\begin{enumerate}
	\item For short-term forecasts, phenomenological time-series approaches are hard to beat, whereas longer-term forecasts require accounting for the influence of slow processes such as evolutionary and ecological selection as well as dispersal.
	\item Different kinds of data reflect the operation of different processes: longitudinal data capture autocorrelation and fast responses of current assemblages to interannual environmental variation, while data spanning spatial gradients capture the long-term outcome of interactions between fast and slow processes.  Whether predictive models should be trained using longitudinal or spatial data sets, or both, depends on the time-scale of the desired forecast.
	\item A key challenge for future research is determining the rate at which slow processes begin to influence dynamics.
\end{enumerate}

@@

The model features Lotka-Volterra competitive interactions among plants within sites that are arrayed along an elevation and temperature gradient. Composition varies along the gradient because of a trade-off between growth rate and cold tolerance: cold sites are dominated by slow-growing species that can tolerate low temperatures, while warm sites are dominated by  fast-growing species that are cold intolerant. Multiple species can coexist within sites because all species experience stronger competition from conspecifics than from heterospecifics. Sites are linked by dispersal: a specified fraction of each species' offspring leaves the site where they were produced and reaches all other sites with equal probability. We provide a more detailed description of the simulation model in SI Appendix \ref{metacomm}. \new{Parameter values for the simulations described here in the main text are shown in Table \ref{tab:comm-turn-pars}.}

We first simulated a baseline period with variable but stationary temperature, followed by a period of rapid temperature increase, and then a final period of stationary temperature. Interannual variation in temperature is the same at all sites, but mean temperature varies among sites. All sites experienced the same absolute increase in mean temperature. We focused on the biomass dynamics of one focal species that dominated the central site during the baseline period. \new{We report results from one simulation run; results were qualitatively consistent for replicate runs.}

@@

\new{When we repeated the simulation with a continuous, nonstationary temperature increase, we see a qualitatively similar shift in weights with increasing forecast horizon from the temporal to spatial statistical model (SI Fig. \ref{fig:community-forecast-species-nonstationary}). In this case, the forecast from the temporal statistical model is not as skillful in the near-term forecast horizon, because the model does not account for the temperature trend during the model fitting period. Separating the effect of annual temperature deviations from the temperature trend would distinguish between short and long-term patterns, much as our temporal and spatial statistical models do in the simulation with stationary temperature periods. }

@@

Our model assumes that fecundity is temperature dependent, and different genotypes have different temperature optima (Fig. \ref{fig:evo_evo_spatial_temporal_models}A). All seeds germinate every year, preventing a seedbank from developing. The model describes how the local density of each genotype changes between years, which depends on temperature and genotype densities in the previous year. Transient temporal dynamics are computed directly from the model; these dynamics are the basis for the temporal statistical model. To create a spatial gradient, we simulated the equilibrium density of each genotype in a series of local populations experiencing different mean temperatures. The pattern of equilibrium densities across the mean annual temperature gradient is the basis for our spatial statistical model: cold sites will be dominated by the cold-adapted homozygous genotype, warm sites will be dominated by the heat-adapted homozygous genotype, and intermediate sites will be dominated by the heterozygous genotype (Fig. \ref{fig:evo_evo_spatial_temporal_models}B). The full description of the eco-evolutionary simulation model is provided in SI Appendix \ref{eco-evo}, and \new{parameter values for simulations described here are shown in Table \ref{tab:eco-evo-pars}. As in the first case study, we report results from just one simulation run, but results were qualitatively consistent for replicate runs.}

@@

Ecological forecasts are typically made using either a space-for-time substitution approach based on models fit to spatial data or using dynamic models fit to time-series data. \new{Empirical studies show that the environment-response relationships detected by these approaches frequently differ in magnitude and even sign \citep{lauenroth_long-term_1992,Oedekoven2017,Amburgey2018,kleinhesselink_response_2018}. Our simulations illustrate how such differences may arise and then lead to} very different predictions about the future state of ecological systems. Which approach provides the most accurate forecasts \new{likely} depends on the forecast-horizon. In our simulations, time-series approaches performed best for short forecast horizons, whereas models based on spatial data made more accurate forecasts at long horizons. In addition, our simulations demonstrate extended transitional periods during which neither the time-series or the spatial approach was effective on its own. The challenge is determining what is ``short-term,'' what is ``long-term,'' and how to handle the many forecasts we need in ecology which fall in between. We have proposed that a weighted combination of the time-series and space-for-time approaches may produce better forecasts at these intermediate forecast horizons.

@@

\new{The idea of combining forecasts with model weights and allowing the weights to shift across the forecast horizon need not be limited to phenomenological models. The same concept could work for any class of models that differ in predictive skill at different forecast horizons. Determining the model weights \emph{a priori} might be easier when models feature explicit processes with characteristic time scales.}

@@

Our \new{simulation studies} have important implications for the emerging field of ecological forecasting. First, they suggest that evaluating model performance at both short and long forecast horizons will be essential as research on forecasting methods accelerates. Second, while single approaches may perform reasonably well for either short or long horizons, skillful predictions at intermediate forecast horizons may require a combination of information from spatial and temporal statistical models. Intermediate time horizons pose challenges in other forecasting contexts as well. Weather forecasts based on regional-scale meteorological models are very effective for forecasting a week to ten days in advance, but then become largely uninformative. Forecasting these intermediate scales has been challenging in meteorology and will likely be challenging in ecology as well. While the recent emphasis on near-term iterative forecasting \citep{dietze_iterative_2018} is the logical and tractable starting point, we also need to build understanding and capacity for forecasting ecological dynamics across all forecast horizons of interest.

@@

\new{To assign species-specific parameter values, the number of species in the metacommunity is specified. Next, each species is assigned an optimal temperature within a specified temperature range by drawing from a uniform distribution. Sensitivity to interspecific competition is then determined as a decreasing function of optimal temperature. Calculations are performed in the script \texttt{SpeciesPoolGen.R}.   }

@@

\begin{figure}[tbp]
	\centering
	\includegraphics[width=0.7 \textwidth] {community_forecast_species_nonstationary.png}
	\caption{(A) Simulated annual temperatures (grey) and expected temperature (black), which was used to make forecasts, at the focal site. \new{In contrast to Fig. \ref{fig:community-forecast-species}, which shows results for a period of warming followed by stationary temperatures, for this simulation we spread the same temperature increase out over the entire simulation with no stationary periods.} (B) Simulated focal species biomass and forecasts from the spatial, temporal and weighted statistical models at the focal site in the metacommunity model. (C) Simulated biomass of the focal species (black) and all other species (grey), and the weight given to the temporal statistical model for focal species biomass (blue). Year 1000 in each panel corresponds to the start of the temperature increase.   }
	\label{fig:community-forecast-species-nonstationary}
\end{figure}

>>>>

\noindent \textbf{Abstract:}   Most phenomenological, statistical models used to generate ecological forecasts take either a time-series approach, based on long-term data from one location, or a space-for-time approach, based on data describing spatial patterns across environmental gradients. However, the magnitude and even the sign of environment-response relationships detected using these two approaches often differs, leading to contrasting predictions about responses to future environmental change. Here we consider how the forecast horizon determines whether more accurate predictions come from the time-series approach, the space-for-time approach, or a combination of the two. We use two simulated case studies to show that forecasts for short and long forecast horizons need to focus on different ecological processes, which are reflected in different kinds of data. In the short-term, dynamics reflect initial conditions and fast processes such as birth and death, and the time-series approach makes the best predictions. In the long-term, dynamics reflect the additional influence of slower processes such as evolutionary and ecological selection, colonization and extinction, which the space-for-time approach can effectively capture. At intermediate time-scales, a weighted average of the two approaches shows promise. However, making this weighted model operational will require new research to predict the rate at which slow processes begin to influence dynamics.

@@

Although both time-series and space-for-time approaches are widely used, there has been little discussion of their advantages and disadvantages for guiding policy decisions or advancing our understanding of ecological dynamics \citep{harris_forecasting_2018,Renwick2018}. Such a discussion is overdue given that these two approaches are likely to make very different predictions about ecological responses to future environmental change. For example, average forage production in the U.S. shortgrass steppe increases rapidly with increasing mean annual precipitation, but production is much less sensitive to interannual variation in precipitation at any one location \citep{lauenroth_long-term_1992}. The space-for-time approach would predict large changes in production in response to a future change in precipitation, while the time-series approach would predict smaller changes. Bird species' abundances in the United Kindgom also show different responses to spatial vs. temporal variation in weather covariates. Not only are abundances typically more sensitive to spatial variation in the covariates, but in many cases the temporal response takes the opposite sign \citep{Oedekoven2017}. Whenever the sign or magnitude of relationships based on time-series and space-for-time approaches differ, so will the resulting forecasts.

@@

Here we use two simulated case studies to 1) demonstrate why time-series and space-for-time approaches can make different predictions, 2) propose that the best model-building approaches for ecological forecasting 
may depend on the time horizon of the forecast, and 3) explore how time-series and space-for-time approaches might be 
combined via weighted averaging to make better forecasts at intermediate time scales. The first case study
focuses on how interspecific interactions affect the population dynamics of a focal species, and the second focuses on
an eco-evolutionary scenario. Our analyses show that: 
\begin{enumerate}
	\item For short-term forecasts, phenomenological time-series approaches are hard to beat, whereas longer-term forecasts require accounting for the influence of slow processes such as evolutionary and ecological selection as well as dispersal.
	\item Different kinds of data reflect the operation of different processes: longitudinal data capture autocorrelation and fast responses of current assemblages to interannual environmental variation, while data spanning spatial gradients capture the long-term outcome of interactions between fast and slow processes.  Whether predictive models should be trained using longitudinal or spatial data sets, or both, depends on the time-scale of the desired forecast.
	\item A key challenge for future research is determining the rate at which slow processes begin to influence dynamics.
\end{enumerate}

@@

The model features Lotka-Volterra competitive interactions among plants within sites that are arrayed along an elevation and temperature gradient. Composition varies along the gradient because of a trade-off between growth rate and cold tolerance: cold sites are dominated by slow-growing species that can tolerate low temperatures, while warm sites are dominated by  fast-growing species that are cold intolerant. Multiple species can coexist within sites because all species experience stronger competition from conspecifics than from heterospecifics. Sites are linked by dispersal: a specified fraction of each species' offspring leaves the site where they were produced and reaches all other sites with equal probability. We provide a more detailed description of the simulation model in SI Appendix \ref{metacomm}. Parameter values for the simulations described here in the main text are shown in Table \ref{tab:comm-turn-pars}.

We first simulated a baseline period with variable but stationary temperature, followed by a period of rapid temperature increase, and then a final period of stationary temperature. Interannual variation in temperature is the same at all sites, but mean temperature varies among sites. All sites experienced the same absolute increase in mean temperature. We focused on the biomass dynamics of one focal species that dominated the central site during the baseline period. We report results from one simulation run; results were qualitatively consistent for replicate runs.

@@

When we repeated the simulation with a continuous, nonstationary temperature increase, we see a qualitatively similar shift in weights with increasing forecast horizon from the temporal to spatial statistical model (SI Fig. \ref{fig:community-forecast-species-nonstationary}). In this case, the forecast from the temporal statistical model is not as skillful in the near-term forecast horizon, because the model does not account for the temperature trend during the model fitting period. Separating the effect of annual temperature deviations from the temperature trend would distinguish between short and long-term patterns, much as our temporal and spatial statistical models do in the simulation with stationary temperature periods.

@@

Our model assumes that fecundity is temperature dependent, and different genotypes have different temperature optima (Fig. \ref{fig:evo_evo_spatial_temporal_models}A). All seeds germinate every year, preventing a seedbank from developing. The model describes how the local density of each genotype changes between years, which depends on temperature and genotype densities in the previous year. Transient temporal dynamics are computed directly from the model; these dynamics are the basis for the temporal statistical model. To create a spatial gradient, we simulated the equilibrium density of each genotype in a series of local populations experiencing different mean temperatures. The pattern of equilibrium densities across the mean annual temperature gradient is the basis for our spatial statistical model: cold sites will be dominated by the cold-adapted homozygous genotype, warm sites will be dominated by the heat-adapted homozygous genotype, and intermediate sites will be dominated by the heterozygous genotype (Fig. \ref{fig:evo_evo_spatial_temporal_models}B). The full description of the eco-evolutionary simulation model is provided in SI Appendix \ref{eco-evo}, and parameter values for simulations described here are shown in Table \ref{tab:eco-evo-pars}. As in the first case study, we report results from just one simulation run, but results were qualitatively consistent for replicate runs.

@@

Ecological forecasts are typically made using either a space-for-time substitution approach based on models fit to spatial data or using dynamic models fit to time-series data. Empirical studies show that the environment-response relationships detected by these approaches frequently differ in magnitude and even sign \citep{lauenroth_long-term_1992,Oedekoven2017,Amburgey2018,kleinhesselink_response_2018}. Our simulations illustrate how such differences may arise and then lead to very different predictions about the future state of ecological systems. Which approach provides the most accurate forecasts likely depends on the forecast-horizon. In our simulations, time-series approaches performed best for short forecast horizons, whereas models based on spatial data made more accurate forecasts at long horizons. In addition, our simulations demonstrate extended transitional periods during which neither the time-series or the spatial approach was effective on its own. The challenge is determining what is ``short-term,'' what is ``long-term,'' and how to handle the many forecasts we need in ecology which fall in between. We have proposed that a weighted combination of the time-series and space-for-time approaches may produce better forecasts at these intermediate forecast horizons.

@@

The idea of combining forecasts with model weights and allowing the weights to shift across the forecast horizon need not be limited to phenomenological models. The same concept could work for any class of models that differ in predictive skill at different forecast horizons. Determining the model weights \emph{a priori} might be easier when models feature explicit processes with characteristic time scales.

@@

Our simulation studies have important implications for the emerging field of ecological forecasting. First, they suggest that evaluating model performance at both short and long forecast horizons will be essential as research on forecasting methods accelerates. Second, while single approaches may perform reasonably well for either short or long horizons, skillful predictions at intermediate forecast horizons may require a combination of information from spatial and temporal statistical models. Intermediate time horizons pose challenges in other forecasting contexts as well. Weather forecasts based on regional-scale meteorological models are very effective for forecasting a week to ten days in advance, but then become largely uninformative. Forecasting these intermediate scales has been challenging in meteorology and will likely be challenging in ecology as well. While the recent emphasis on near-term iterative forecasting \citep{dietze_iterative_2018} is the logical and tractable starting point, we also need to build understanding and capacity for forecasting ecological dynamics across all forecast horizons of interest.

@@

To assign species-specific parameter values, the number of species in the metacommunity is specified. Next, each species is assigned an optimal temperature within a specified temperature range by drawing from a uniform distribution. Sensitivity to interspecific competition is then determined as a decreasing function of optimal temperature. Calculations are performed in the script \texttt{SpeciesPoolGen.R}.

@@

\begin{figure}[tbp]
	\centering
	\includegraphics[width=0.7 \textwidth] {community_forecast_species_nonstationary.png}
	\caption{(A) Simulated annual temperatures (grey) and expected temperature (black), which was used to make forecasts, at the focal site. In contrast to Fig. \ref{fig:community-forecast-species}, which shows results for a period of warming followed by stationary temperatures, for this simulation we spread the same temperature increase out over the entire simulation with no stationary periods. (B) Simulated focal species biomass and forecasts from the spatial, temporal and weighted statistical models at the focal site in the metacommunity model. (C) Simulated biomass of the focal species (black) and all other species (grey), and the weight given to the temporal statistical model for focal species biomass (blue). Year 1000 in each panel corresponds to the start of the temperature increase.   }
	\label{fig:community-forecast-species-nonstationary}
\end{figure}
